Aligning marine species range data to better serve science and conservation

### 2016-08-17 ms_1_intro.Rmd
sent to BH, JA, CS

___sentence about SAUP - how can we rephrase the "two global-scale repositories" sentence to exclude it? "broad taxonomic coverage"?  something about not focusing just on economically important species? (is that a valid assumption about SAUP)___

@@

__Neither IUCN nor AquaMaps claims a lock on the "truth", and the small overlap in mapped species means each provides value in taxonomic and geographic coverage.  Method-driven differences in commission and omission errors produce clear and significant disagreement in species range descriptions between AquaMaps and IUCN datasets.__

Neither dataset claims to represent the "truth" of a species' spatial distribution, but rather each presents an understanding of the "truth" - one relying primarily on expert opinion and the other on model predictions, each method with its own challenges and advantages.  Uncertainties inherent in method and intent inevitably drive differences in the range predictions made by each dataset; significant differences could indicate dramatically different understandings of our marine ecosystems, with significant implications for policy and conservation recommendations.  Inaccurate indications of presence or absence could lead to ineffective marine reserve systems and management plans [@rondinini_tradeoffs_2006; @jetz_ecological_2008].  Importantly, biases in taxonomic or spatial coverage within a dataset could shift management and conservation actions away from places or species that are most in need.

@@

As understanding of a given species improves, we expect that range maps, regardless of method, would become more accurate in predicting both species presence and absence.  Close alignment between two independent range maps can therefore indicate higher confidence in species presence or absence than a single map.  For the `r format(n_spp_both, big.mark = ',')` species (`r p_both_all`% of total) mapped in both datasets, we examine how well the maps align in both distribution and overall area; from the results we determine several issues that can introduce commission errors, and suggest methods to improve confidence in species range predictions.

@@

While we cannot recommmend one data set as more accurate than the other, we suggest methods to improve confidence in species range predictions to better inform conservation management and policy decisions.

___acknowledge or debunk other global data sets - e.g. SAUP?___

>>>>

___sentence about SAUP - how can we rephrase the "two global-scale repositories" sentence to exclude it? "broad taxonomic coverage"?  something about not focusing just on economically important species? (is that a valid assumption about SAUP) - consider other datasets as well: NatureServe, BOTW, etc___

@@

Neither dataset claims to represent the "truth" of a species' spatial distribution, but rather each offers an understanding of the "truth" - one relying primarily on expert opinion and the other on model predictions.  Each method presents its own challenges and advantages, and the small overlap in mapped species means each dataset provides value in taxonomic and geographic coverage.  Importantly, biases in taxonomic or spatial coverage within a dataset could shift management and conservation actions away from places or species that are most in need.

Uncertainties inherent in method and intent inevitably drive differences in the range predictions made by each dataset: geographic range data such as IUCN range maps frequently introduce commission errors (inaccurate indication of presence), while species distribution models such as AquaMaps will likely introduce fewer commission errors but more omission errors (inaccurate indications of absence) [@rondinini_tradeoffs_2006].  Each type of error bears different implications for conservation goals: commission errors can result in prioritizing areas not relevant to conservation goals, while omission errors may result in protected area networks that fail to include important habitat and range [@rondinini_tradeoffs_2006; @jetz_ecological_2008].

@@

As understanding of a given species improves, we expect that range maps, regardless of method, would become more accurate in predicting species presence and absence.  Close alignment between two independent range maps can therefore indicate higher confidence in species presence or absence than a single map.  For the `r format(n_spp_both, big.mark = ',')` species (`r p_both_all`% of total) mapped in both datasets, we examine how well the maps align in both distribution and overall area; from the results we determine methodological issues that can introduce commission errors, and suggest methods to improve confidence in species range predictions.

@@

Each of these datasets provides unique value in taxonomic and geographic coverage; while we have identified some potential issues in each, we cannot simply recommmend one data set over the other.  Instead, we recommend simple methods to reduce incidence of commission errors in each dataset, improving confidence in species range predictions to better inform conservation management and policy decisions.

### 2016-08-17 ms_2_methods.Rmd
sent to BH, JA, CS

# Methods

_About the datasets__:

The IUCN publishes species range maps based on expert input of spatial boundaries of a given species' "limits of distribution" [@iucn_iucn_2015] - essentially a refined extent of occurrence (EOO), based on observation records and informed by expert understanding of species' range and habitat preferences.  In contrast, AquaMaps models species distribution based on environmental preferences (e.g., temperature, depth, salinity) deduced from occurrence records, published species databases such as FishBase, and expert knowledge.  The AquaMaps model overlays these environmental preferences atop a map of environmental attributes on a 0.5 degree grid, creating a global map of suitable habitat.  The model masks habitat to certain FAO Major Fishing Areas as a rough cut of geographic range, resulting in a "probability of occurrence" for each species.  Of the resulting maps, 1296 (5.7%) have been further refined under expert review [@kaschner_mapping_2006; @ready_predicting_2010].  Studies using AquaMaps data frequently define "presence" by applying a probability threshold, e.g., previously, the Ocean Health Index defined presence as 40% or greater probability of occurrence [@halpern_index_2012; @halpern_patterns_2015].  In the 2016 OHI assessment, the presence threshold was eliminated, to better approximate the "extent of occurrence" generally indicated by IUCN maps, since the two datasets are used together.

___short paragraph about other datasets we didn't include, i.e. SAUP? BOTW?___

@@

__Comparison of paired maps__:
Using genus and species binomials as a matching key, we selected the "paired map" species - the subset of marine species that have range maps in both IUCN and AquaMaps current native distribution.  _We used the taxize package [@REF] in R [@REF] to standardize species names and synonyms; for species with separate subpopulation maps in IUCN, we combined all subpopulations to create a single global population._   For each of these paired map species, we determined species presence within each spatial cell for each dataset, using the same criteria outlined above.

Overlaying paired distribution maps for each species, we defined and calculated _distribution alignment_ and _area ratio_ and plotted these in Fig. 2A:

@@

_Distribution alignment uses overlapping predictions of presence as means of identifying areas of high confidence.  Area ratio provides a proxy for frequency of commission and/or omission errors._

Dividing the plot into quadrants based on mean values along each axis, we roughly categorized four qualities of alignment, and for each taxonomic group, based on IUCN taxonomic groupings, determined the proportion of paired map species that fell into each quality category (2B).

_Based on the results of the map alignment analysis, we chose to examine the effect of explicitly restricting IUCN ranges to depths based on species' life histories.  We selected corals as a case study, due to their foundational role in biodiverse habitats, as well as their lack of mobility and general reliance on photosynthesis._  Using the IUCN's API [REF] to find life history information for all IUCN-mapped corals, we found that none of the paired-map corals is described as occurring deeper than 200 m; 94% are noted as being limited to 50 m or shallower.  We created a 200 m bathymetry raster from Natural Earth's 200 m bathymetry polygon [@natural_earth_XXX] ___(Fig. SXXX)___ and masked our IUCN coral rasters to exclude presence below 200 m.  The resulting maps were again compared to the AquaMaps rasters; ___Fig. XXXX___ plots shifts in coral species alignment due to this depth clipping.

___To identify data-poor species, we used the robis??? and rgbif??? packages in R to identify known occurrences of each species; occurrences were averaged between the two bases as a proxy for data richness.___

**Methods for Ocean Health Index (OHI) case study:**
The Species component of the OHI Biodiversity goal (SPP) uses IUCN Red List extinction risk as a proxy for a species' health.  Each species is assigned a risk score based on IUCN Red List extinction risk category, from Least Concern (0.0) to Extinct (1.0) (Data Deficient and unassessed species are excluded).  Species range is determined from its IUCN range map, where available, or its AquaMaps range map if an IUCN map is unavailable.  For each assessed OHI region, an area-weighted mean extinction risk across all species is calculated, and rescaled to calculate a status score from 0 (mean extinction risk = 0.75) to 100 (mean extinction risk = 0) [@halpern_index_2012; @halpern_patterns_2015].

_To highlight the differences in the datasets, we recalculated the OHI SPP status values using only one complete dataset at a time.  For the AquaMaps dataset, we used a presence threshold of 0% to align with the OHI 2016 assessment. Fig. 3 compares these single-dataset scores to the 2016 OHI SPP status values.  While the full OHI 2016 assessment includes the Birds of the World dataset [@REF???] in addition to IUCN and AquaMaps, we omitted it from our calculations for simplicity._

~To highlight the differences in the datasets, we selected the subset of species whose ranges are represented in both datasets, eliminating species found in only one or the other dataset.  In the original OHI SPP analysis, IUCN data was used to determine ranges for these overlapping species; we calculated all regional SPP status scores for just these species as our baseline.  We then recalculated the scores using AquaMaps ranges at a 40% presence threshold (as used in the original OHI analysis) and a 0% threshold (to better simulate the "limits of distribution" defined by IUCN maps) and compared them to our IUCN baseline.~

~We then modified the original OHI source code to recalculate the 2015 SPP status scores for all regions and all species to examine the effects of changing the AquaMaps presence threshold from 40% to 0% (all range included), and the effects of changing the data source for the overlapping species from IUCN to AquaMaps.  We calculated the differences between these results and the published SPP status scores for 2015.~

**Methods for MPA Gap Analysis case study:** 
To assess the effectiveness of MPAs in protecting biodiversity, Klein et al. [-@klein_shortfalls_2015] compared the coverage of the global MPA network presented by the World Database on Protected Areas (WDPA) [@iucn_world_2014] to the species ranges described in the 2014 AquaMaps dataset [@aquamaps_XXX].  For the primary analysis, the researchers defined species "presence" as 50% or greater probability of occurrence.

To reconstruct the primary analysis, we selected the subset of protected areas from the 2014 WDPA dataset classified as IUCN protected area management categories I-IV and spatially covering a marine area. The WDPA polygons and marine polygons were rasterized to 0.01° and then aggregated to AquaMaps native 0.5° cells, to calculate proportion of marine protected area within each cell.  After verifying our results using the 2014 AquaMaps dataset, we updated the analysis to use the 2015 AquaMaps dataset, at a presence threshold of 50% (to compare to Klein et al. directly) and 0% (to better compare with IUCN spatial data).  To analyze MPA coverage against IUCN spatial data, we extracted IUCN polygon weights per 0.5° cell for each species and compared against the protected area raster.  Finally, we combined AquaMaps data (at 0% threshold) and IUCN data, using AquaMaps for the `r format(n_spp_both, big.mark = ',')` overlapping species and again compared against the protected area raster.

>>>>

# Methods and Analysis

## About the datasets

The IUCN publishes species range maps developed by species experts.  These experts outline spatial boundaries of a given species' "limits of distribution" [@iucn_iucn_2015], based on observation records and informed by expert understanding of species' range and habitat preferences.  ___Fig. SXXX___

In contrast, AquaMaps models species distribution based on environmental preferences, such as temperature, depth, and salinity, deduced from occurrence records, published species databases such as FishBase, and expert knowledge.  The AquaMaps model overlays these environmental preferences atop a map of environmental attributes on a global 0.5° grid to determine suitable habitat, resulting in a "probability of occurrence" for each species.  Of these, 1296 (5.7%) have been further refined through an expert review process [@kaschner_mapping_2006; @ready_predicting_2010]. ___Fig. SXXX___

___OMIT: Studies using AquaMaps data frequently define "presence" by applying a probability threshold, e.g., previously, the Ocean Health Index defined presence as 40% or greater probability of occurrence [@halpern_index_2012; @halpern_patterns_2015].  In the 2016 OHI assessment, the presence threshold was eliminated, to better approximate the "extent of occurrence" generally indicated by IUCN maps, since the two datasets are used together.___

@@

__Comparison of paired maps__:
Using genus and species binomials as a matching key, we identified "paired map" species - the subset of marine species that have range maps in both IUCN and AquaMaps current native distribution.  We used the taxize package [@REF] in R [@REF] to standardize species names and synonyms; for species with separate subpopulation maps in IUCN, we combined all subpopulations to create a single global population.   For each of these paired map species, we determined species presence within each spatial cell for each dataset using the same criteria outlined above.

Overlaying paired distribution maps for each species, we defined and calculated _distribution alignment_ $\alpha_{dist}$ and _area ratio_ $\alpha_{area}$:

@@

Distribution alignment uses overlapping predictions of presence as means of identifying areas of higher confidence.  Area ratio provides a proxy for frequency of commission and/or omission errors.

To examine errors of commission related to depth, we selected corals as a case study due to their importance in supporting biodiversity as well as their dependence on photosynthesis.  We created a 200 m bathymetry raster from Natural Earth's 200 m bathymetry polygon [@natural_earth_XXX] ___(Fig. SXXX)___ and masked our IUCN coral rasters to identify mapped coral presence below 200 m.  The resulting maps were again compared to the AquaMaps rasters to examine distribution alignment and area ratio.

To identify data-poor species, we used the ___robis???___ and ___rgbif???___ packages [@REFS] in R to identify known occurrences of each species; occurrences were averaged between the two occurrence databases as a proxy for data richness.

## Methods for Ocean Health Index (OHI) case study

The global Ocean Health Index (OHI) [@halpern_index_2012; @halpern_patterns_2015], a composite index comprising ten sustainable benefits provided by a healthy ocean, uses species spatial distribution data to calculate biodiversity status for each of the world's 221 exclusive economic zones.

The Species component of the OHI Biodiversity goal (abbreviated as SPP) uses IUCN Red List extinction risk as a proxy for a species' health.  Each species is assigned a risk score based on IUCN Red List extinction risk category, from Least Concern (0.0) to Extinct (1.0) (Data Deficient and unassessed species are excluded).  Species range is determined from its IUCN range map, where available, or its AquaMaps range map if an IUCN map is unavailable.  For each assessed OHI region, an area-weighted mean extinction risk across all species is calculated, and rescaled to calculate a status score from 0 (mean extinction risk = 0.75) to 100 (mean extinction risk = 0) [@halpern_index_2012; @halpern_patterns_2015].

_To highlight the differences in the datasets, we recalculated the OHI SPP status values using only one complete dataset at a time, and compared to the values resulting from the combined datasets.  For the AquaMaps dataset, we used a "probability of occurrence" threshold of 0% to determine presence, to align with the OHI 2016 assessment._

## Methods for MPA Gap Analysis case study
To assess the effectiveness of MPAs in protecting biodiversity, Klein et al. [-@klein_shortfalls_2015] compared the coverage of the global MPA network presented by the World Database on Protected Areas (WDPA) [@iucn_world_2014] to the species ranges described in the 2014 AquaMaps dataset [@aquamaps_XXX].  For the primary analysis, the researchers defined species presence as 50% or greater probability of occurrence.

To reconstruct the primary analysis, we selected the subset of protected areas from the 2014 WDPA dataset classified as IUCN protected area management categories I-IV and spatially overlapping a marine area. The WDPA polygons and marine polygons were rasterized to 0.01° and then aggregated to AquaMaps native 0.5° cells, to calculate proportion of marine protected area within each cell.  After verifying our results using the 2014 AquaMaps dataset, we updated the analysis to use the 2015 AquaMaps dataset, at a presence threshold of 50% (to compare to Klein et al. directly) and 0% (to better compare with IUCN spatial data).  To analyze MPA coverage against IUCN spatial data, we extracted IUCN polygon weights per 0.5° cell for each species and compared against the protected area raster.  Finally, we combined AquaMaps data (at 0% threshold) and IUCN data, using AquaMaps for the `r format(n_spp_both, big.mark = ',')` overlapping species and again compared against the protected area raster.
